Contributions of adaptation and purifying selection to SARS-CoV-2 evolution

### 2022-10-30 manuscript/response_to_reviewers.tex

\editor{

Three reviewers have seen the paper, and I completely agree with all three that this is an excellent paper that uses a robust & rapid approach to obtain meaningful insights about SARS-CoV-2 evolution. The paper could be published in its current form but would benefit from very minor corrections suggested by the reviewers.
I also add a very small concern of my own: citing from the paper "...remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations". This is a little ad-hoc, how much would the results change if changing this? How many sequences are removed this way? To emphasize, I don't think any in-depth analysis is necessary here, just a few sentences on why this cutoff is used.
Looking forward to seeing this paper published,
Adi
}

\response{Thanks a lot for handling the paper, the constructive comments, and the rapid turn-around.

You are completely right that these cutoffs are heuristic and I have checked that the results are robust.

The number of outliers removed varies from clade to clade as the clades, but never exceeds 1\% with the standard parameters.
Changing these parameters (using 4 instead of 2 sigma of the Poisson distribution, 6 instead or 3 mutations offset which reduces outlier removal by 90\%) changes the rate estimates by a less than one percent (3\% at most for clade 20H).
The relative insensitivity to these outliers is due to the fact that we don't build a tree and a few outliers don't distort the overall picture.
Similarly, making the QC filter less stringent (using 80 instead of 30) doesn't change rate estimate by more than 2\% on average. The biggest effect of the QC filter is filtering for completeness.
I have added a few sentences to highlight this.

}

@@

\response{Thank you very much.}

\refa{My major comment is related to the author's acknowledgement that, ``Nextstrain clades tend to be defined by long branches leading to a large polytomy. It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.'' As the author describes elsewhere, there are many potential reasons for the appearance of these long branches -- cryptic transfer, prolonged infection within a clinical population, reverse zoonosis. However, as the author states, none of these factors have been conclusively demonstrated for specific VOC emergence much less in general and it is possible that the mutations placed on these long branches emerged within hosts part of acute transmission chains belonging to robustly sequenced populations. Even if this is the case, while the sequencing efforts throughout the pandemic have been unprecedented, they are nonuniform and sequencing efforts (and reporting biases) are influenced by the emergence of mutations of interest. I believe it's likely the relative number of infections in the period immediately ancestral to the last common ancestor is underestimated relative to that of the period immediately following for VOC clades which may impact the estimation of relative mutation rates. This is exacerbated by changing public behavior (less likely to engage in social settings after VOC emergence). This effect may be more pronounced for nonsynonymous events than synonymous events due to the nature of the tree reconstruction.}

\response{This is an interesting thought. However, I am not 100\% sure I understand how this differential ascertainment would affect the accumulation of mutations or inferences of their rate.}

@@

\refb{
This is a very nice study looking at the rate of evolution in SARS-CoV-2. Whereas other studies have looked at similar questions using increasingly complicated phylogenetic methods, Richard uses simple robust model that allows him to easily analyse huge numbers of sequences. Because of the wealth of SARS-CoV-2 data even a very simple model will illuminate the big patterns. This is an important reminder to all of us: that a model need only be complicated enough to answer the questions at hand. While hard numbers aren't given I assume that millions of genomes went into the analyses presented here. (The Tay and Hill papers cited here could only use hundreds of sequences). One small caveat of the approach is that it isn't suited to all scenarios, but is especially suited to clades with explosive exponential growth (this is also mentioned in the manuscript).

The main conclusions are:
\begin{enumerate}
 \item Synonymous mutations are mostly neutral
 \item The rate of adaptation clearly slowed over the first few months of the pandemic
 \item The process that gave rise to VOCs was decidedly different than normal human-human transmission
 \item The rate of adaptation and the slowdown are highly variable across the genome
\end{enumerate}
Perhaps most interesting to me is that the slowdown in the within-clade evolutionary rate is not seen in Spike, which is most important for infectivity and immune escape. However, Spike still appears to be under quite strong evolutionary constraint, as can be seen from it consistently having one of the slowest within-clade evolutionary rates (Fig. S1). None of these are particularly surprising conclusions, but this is the most elegant way I've seen of reaching them.

Overall I think the results are very interesting, the manuscript is well-written and the figures are good. I only have a few small comments.}

\response{Thank you very much for this accurate summary and the generous remarks!}

@@

\refb{1. Can this truly be said to be a tree-free approach? Presumably some phylogenies are calculated in order to assign sequences to Nextstrain clades? However, since this is continuously done and clade assignments are public it can be argued that this initial part of the analysis is "free".}

\response{The reviewer is correct. Sequences are assigned to clades by placing them on a reference tree.
Though this would also be possible using a constellation of mutations or some other sort of clustering. If some clades were the result of recombination (as we soon might experience), the approach would work exactly the same way. So I think it is fair to say that most the results presented do not make explicit use of phylogenies, though some implicit usage of phylogenetic concepts (clade and lineage assignment) is present. I chose to not discuss this point at length.  }

@@

\refb{2. Does filtering out all sequences in a clade that don't have the full set of mutations relative to the reference mean you are excluding sequences with reversions to the reference? Do we know how often reversions occur? I don't think they are common, although they have been seen for Omicron (also mentioned here).}

\response{This is an important point (and I tripped over issues related to this when doing the analysis of fitness costs). Importantly, the requirement for a sequence to have the complete set of mutations refers to the clade defining mutations. Omicon clades (22A/BA.4, 22B/BA.5) have reversions and the reverted positions are not part of the clade defining mutations (their genotype is ancestral). Ignoring sequences that don't have the full set of clade defining mutation will ignore sequences that have additional reversions at clade defining positions. Until summer 2022, this was not much of a concern and most reversions within clades are clear QC issues. For the analysis of fitness costs and purifying selection, all sequences that pass the qualitaty thresholds are included and reversions are counted.
I have made this clearer in the manuscript.}

@@

\refc{I would be interested to have a definition (or a citation) of what constitutes a nextstrain clade -- I assume it is phylogenetic, but just a quick summation of what the rules are to define it would be helpful, especially since it is different to the more commonly used pango lineage system.}

\response{Yes, Nextstrain clades are phylogenetically defined and assigned using Nextclade. I have added a reference to a blog post on nextstrain.org that outlines the clade definitions.}

\refc{Paragraph starting ``In other cases, notable subclades…'' page 3 line 55 in the second column -- I don't follow the logic here of increased divergence ``generating an offset to diversity''. What is diversity if not within-clade divergence?}

\response{Yes, this was poorly worded. What I mean to say is that when there are many lineages, time since emergence can estimated accurately from the divergence of many independent evolutionary trajectories. When there is a single subclade or a few large subclades, estimates of emergence times are more noisy. I have made an effort to reword this. }

@@

\refc{Figure 2 legend: ``Genotypes in D refer to sequences the all and only the mutations indicated'' -- are there some words missing here?}

\response{Yes. Thanks. Another reviewer picked this up as well and this is now fixed.}

\refc{I would find it helpful to have the evolutionary rates in substitutions/site/year as well as the rate per year in figure 3 (in the overall rate annotation)}

\response{Thank you for this suggestion, I have added this for panels A and B. If give amino acid and synonynmous rates 'per codon' to avoid subtleties counting synonymous sites (2-fold, 4-fold, etc). }

### 2022-10-30 manuscript/variant_rates.tex
work on revision

For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of clade defining mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.

@@

When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and Supp.~Fig.~\ref{fig:fitness_landscape}), the most constrained regions are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
\emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
Variation at 3rd positions is common and comparable between genes.
Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35 (see Fig.~\ref{fig:fitness_landscape}).
Systematic difference in purifying selection on the amino acid sequence of different reading frames have also been reported by \citet{rochman_ongoing_2021}.

@@

Most of this apparent initial acceleration of within-variant evolution ceased in mid-2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
By that time, the number of non-synonymous differences relative to the root of the tree was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
So why would the rate of adaptive evolution slow down?
Maybe it is just chance, after all it is only clades 19A and 19B where the rate is substantially higher.
Another possible explanation could be \emph{diminishing returns} epistasis.
Viruses with the S:D614G mutation showed faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
The virus might therefore be operating closer to the maximal capacity at which cells can produce virions, reducing the scope for further optimization.
Other mutations then have smaller benefits, rise in frequency more slowly, and the effect of adaptation does not manifest itself over the life time of the clades studied here.
Such diminishing returns epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021,rochman_epistasis_2022}.
Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate but had selective origin.

>>>>

For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of clade defining mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
Results are insensitive to the stringency of this filtering.
For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations. These outliers are never more than 0.5\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.

@@

When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and Supp.~Fig.~\ref{fig:fitness_landscape}), the most constrained regions are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
\emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
Variation at 3rd positions is common and comparable between genes.
Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35 (see Fig.~\ref{fig:fitness_landscape}).
Systematic difference in purifying selection on the amino acid sequence of different open reading frames have also been reported by \citet{rochman_ongoing_2021}.

@@

Most of this apparent initial acceleration of within-variant evolution ceased in mid-2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
By that time, the number of non-synonymous differences relative to the root of the tree was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
So why would the rate of adaptive evolution slow down?
Maybe it is just chance, after all it is only clades 19A and 19B where the rate is substantially higher.
Another possible explanation could be \emph{diminishing returns} epistasis.
Viruses with the S:D614G mutation showed faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
The virus might therefore be operating closer to the maximal capacity at which cells can produce virions, reducing the scope for further optimization.
Other mutations then have smaller benefits, rise in frequency more slowly, and the effect of adaptation does not manifest itself over the life time of the clades studied here.
Such diminishing returns epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021,rochman_epistasis_2022,martin_selection_2022}.
Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate but had selective origin.
